Pangenome graph layout by Path-Guided Stochastic Gradient Descent

### 2023-09-20 manuscript/main.tex
cosmetics

    \address{
        $^1$Quantitative Biology Center (QBiC), University of Tübingen, Tübingen 72076, Germany \\
        $^2$Biomedical Data Science, Department of Computer Science, University of Tübingen, Tübingen 72076, Germany \\
        $^3$Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, TN 38163, USA \\
        $^4$Genomics Research Centre, Human Technopole, Milan 20157, Italy \\
        $^5$Department of Computer Engineering, School of Computation, Information and Technology (CIT), Technical University of Munich, Munich 80333, Germany \\
        $^6$School of Electrical and Computer Engineering, Cornell University, Ithaca, NY 14853, USA \\
        $^7$Computomics GmbH, Eisenbahnstr. 1, 72072 Tübingen, Baden-Württemberg, Germany \\
    }

@@

	\FIXME{@Jiajie + Niklas: Maybe address our SMT issue with your cache locality as a potential solution. Additionally just 2/3 sentences about going GPU?}
	The performance analysis shows that our 2D implementation outperforms \textit{BandageNG} when handling large, complex pangenome graphs.
	While \textit{BandageNG} was not able to deliver a layout of the whole HPRC graph within 1 week, our 2D PG-SGD calculated one within one day.
	%We expect that our implementation will be able to handle the next phase of the HPRC, 24 chromosomal pangenome graphs constructed from 300 individuals, without any problems.
	As especially observed with chromosome 16 evaluation, our implementation is memory-bound.
	The current data structure doesn't consider the memory access pattern.
	Therefore, a CPU cache optimized implementation could make use of the massive parallelized calculations on a GPU.
	Such efforts are currently in progress \url{https://github.com/nsmlzl/odgi/tree/layout_cache_optimized_cpu}.

	PG-SGD can be extended to any number of dimensions.
	It can be seen as a graph embedding algorithm that converts high-dimensional, sparse pangenome graphs into low-dimensional, dense, and continuous vector spaces, while preserving its biologically relevant information.
	This enables the application of machine learning algorithms that use the graph layout for variant detection and classification. 
	Our future research involves leveraging these graph projections to detect structural variants and to identify and correct assembly errors
	Moreover, we are considering extending the algorithm to RNA and protein sequences to support pantranscriptome graphs \citep{sibbesen_haplotype-aware_2023} and panproteome graphs \citep{dabbaghie_panpa:_2023}, respectively.
	%Helpful could be to integrate more layout objectives \citep{ahmed_multicriteria_2021} like neighborhood preservation.

>>>>

    \address{
        $^1$Quantitative Biology Center (QBiC), University of Tübingen, Tübingen 72076, Germany \\
        $^2$Biomedical Data Science, Department of Computer Science, University of Tübingen, Tübingen 72076, Germany \\
        $^3$Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, TN 38163, USA \\
        $^4$Genomics Research Centre, Human Technopole, Milan 20157, Italy \\
        $^5$Department of Computer Engineering, School of Computation, Information and Technology (CIT), Technical University of Munich, Munich 80333, Germany \\
        $^6$School of Electrical and Computer Engineering, Cornell University, Ithaca, NY 14853, USA \\
        $^7$Computomics GmbH, Eisenbahnstr. 1, 72072 Tübingen, Germany \\
    }

@@

	% \FIXME{@Jiajie + Niklas: Maybe address our SMT issue with your cache locality as a potential solution. Additionally just 2/3 sentences about going GPU?}
	The performance analysis shows that our 2D implementation outperforms \textit{BandageNG} when handling large, complex pangenome graphs.
	While \textit{BandageNG} was not able to deliver a layout of the whole HPRC graph within 1 week, our 2D PG-SGD calculated one within one day.
	%We expect that our implementation will be able to handle the next phase of the HPRC, 24 chromosomal pangenome graphs constructed from 300 individuals, without any problems.
	There are some possible optimization approaches for future work to further improve the performance of PG-SGD, making it possible for interactive use. 
	The data structure could be optimized to improve ache performance. Moreover, the high-degree of parallelism could be further exploited by using a GPU.  
	% As especially observed with chromosome 16 evaluation, our implementation is memory-bound.
	% The current data structure doesn't consider the memory access pattern.
	% Therefore, a CPU cache optimized implementation could make use of the massive parallelized calculations on a GPU.
	% Such efforts are currently in progress \url{https://github.com/nsmlzl/odgi/tree/layout_cache_optimized_cpu}.

	PG-SGD can be extended to any number of dimensions.
	It can be seen as a graph embedding algorithm that converts high-dimensional, sparse pangenome graphs into low-dimensional, dense, and continuous vector spaces, while preserving its biologically relevant information.
	This enables the application of machine learning algorithms that use the graph layout for variant detection and classification. 
	Our future research involves leveraging these graph projections to detect structural variants and to identify and correct assembly errors. 
	Moreover, we are considering extending the algorithm to RNA and protein sequences to support pantranscriptome graphs \citep{sibbesen_haplotype-aware_2023} and panproteome graphs \citep{dabbaghie_panpa:_2023}, respectively.
	%Helpful could be to integrate more layout objectives \citep{ahmed_multicriteria_2021} like neighborhood preservation.
